---
title: "PCA Comparison of Discovery and Replication Cohorts"
author: "Graham R Smith"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  github_document: default
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---
  
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

# output batch corrected matrices?
write_output <- FALSE
```
  
  
```{r libraries, echo = F, include = F}

  # 2 cohort pca - derived from qc_swaps.R 
## libraries ----

library(tidyverse)
library(edgeR)
library(limma)
library(splines)
library(reshape2)
library(qvalue)
library(ggsci)
library(cowplot)
library(ggrepel)
library(plotly)
library(gridExtra)
library(bioplotr)
library(umap)


```  
  

Read data back in, remove outlier Patients and Samples and filter genes by expression (cpm > 1 in a fraction > 0.5 of samples)


```{r read, echo = F, include = F}

# old clinfile <- file.path(qcd, "clinical/augmented_clinical_data.csv")
clinfile <- "/data/ngrs2/PSORT/data/David_20_04_2020/combined_clin_w_PASI.csv"

datad <- "/data/ngrs2/PSORT/data/"
annofile <- file.path(datad, "Hs.anno.csv")
dfile <-  file.path(datad, "combined_cnts.rds")

# read annotation 
ad <- read_csv(annofile)
# remove the [Source:HGNC Symbol;Acc:HGNC:48404] stuff
ad$Description <- sub(" \\[Source.*", "", ad$Description)

# read clinical data 
d <- read_csv(clinfile)
# recode ... drug names are different
d <- mutate(d, Drug = recode(Drug, "Adalimumab"="A", "Ustekinumab"="U" ),
            Cohort = recode(Cohort, "Discovery" = "Discovery", "Validation" = "Replication"))
# but use this for plots 
drug_out <- list(A = "ADA", U = "UST")
d <- mutate(d,  Time = recode(Time, "wk00"="0w","wk01"="1w","wk04"="4w","wk12"="12w")) %>% 
              mutate(Time = factor(Time, levels = c("0w","1w","4w","12w")))

d_input <- d 




# get expression data
edata <- read_rds(dfile)
cat("n genes", nrow(edata), "\n")

# filter to tissue 
keep_samp <- d %>% filter (Tissue != "Blood" )

# don't want these
paste0("P", c(60006, 6011, 6042, 6043, 6044, 6045)) %in% keep_samp$Patient_id
# just 6011 is labelled Discovery (they are all present in complete data d )
keep_samp <- filter(keep_samp, ! Patient_id %in% paste0("P", c(60006, 6011, 6042, 6043, 6044, 6045)))

# dont want these either - outliers in blood ... 
to_rm <- c("PSORT-P1-18", "PSORT-P1-71", "PSORT-P1-UHR", "PSORT-P2-UHR-1", "PSORT-P2-UHR-2", "PSORT-P3-06", "PSORT-P3-07", "PSORT-P4-23")
# and this is a duplicate sample for P6001 wk4 blood 
to_rm <- c(to_rm, "PSORT-P1-52")
keep_samp <- filter(keep_samp, ! Sample_id %in% to_rm)


this_data <- DGEList(edata[, keep_samp$Sample_id])
n <- nrow(keep_samp)
keep_gene <- rowSums(cpm(this_data) >= 1) >= n * 0.5
this_data <- DGEList(this_data[keep_gene, ])
this_data <- calcNormFactors(this_data)
#dim(this_data)

voomed_data <- voom(this_data)
# Filter, normalize
#n <- nrow(keep_samp)
#mat <- DGEList(cnts[, tmp$Sample_id])
#keep <- rowSums(cpm(mat) >= 1) >= n / length(unique(tmp$Drug_Time))
#mat <- DGEList(mat[keep, ])
#mat <- calcNormFactors(mat)

```

Numbers of samples and genes

```{r cohort_stats, include = T, echo = F}


cat("after this, number of patients", length(table(keep_samp$Patient_id)), "\n")
cat("number of genes", nrow(this_data), "\n")
cat("number of samples", ncol(this_data), "\n")
cat("distribution of samples over cohorts (D,R):", table(keep_samp$Cohort), "\n")


```

Define a function to do PCA 


```{r pca_fn}
pca_fn <- function(inputvdat) {
  
  logexpr <- rowMeans(inputvdat)
  nk = 5000
  to_keep = head(sort(logexpr,decreasing=TRUE), n = nk)
  topE = inputvdat[names(to_keep),]
  pca = prcomp(t(topE))
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  
  
  ds <- keep_samp
  pd <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], PC3 = pca$x[, 3], PC4 = pca$x[, 4],
                   Patient_id = as.character(ds$Patient_id), 
                   PASI = ds$PASI, Cohort = ds$Cohort, Drug = ds$Drug,
                   Time = ds$Time, Tissue = ds$Tissue)
  
  percentVar <- pca$sdev^2/sum(pca$sdev^2)
  #these_dat <- pd %>% filter(Patient_id %in% int_p) %>% arrange(Patient_id)
  #cat(as.character(these_p),"\n")
  xlim=c(round(min(pd$PC1)),round(max(pd$PC1)))
  ylim=c(round(min(pd$PC2)),round(max(pd$PC2)))
  
  mycols <- pal_d3()(4)
  mycols <- mycols[c(1,3,4,2)]
  names(mycols) <- c("0w","1w","4w","12w")
  
  
  p <- ggplot(data = pd, aes(x = PC1, y = PC2, color = Time, Patient_id = Patient_id)) +
    geom_point(size = 1.5, alpha = 0.5, aes(shape = Tissue)) + 
    xlab(paste0("PC1: ", round(percentVar[1] * 100), "% variance")) + 
    ylab(paste0("PC2: ", round(percentVar[2] * 100), "% variance")) +
    scale_colour_manual(values = mycols) + 
    scale_x_continuous(limits = xlim) + 
    scale_y_continuous(limits = ylim) + ggtitle("Both Cohorts") + 
    guides(size = FALSE, alpha = FALSE) + theme_bw(base_size = 12)
  show(p)
  # show(ggplotly(p))
  

    p <- ggplot(data = pd, aes(x = PC1, y = PC2, color = Time, Patient_id = Patient_id)) +
      geom_point(size = 1.5, alpha = 0.5, aes(shape = Tissue)) + 
      xlab(paste0("PC1: ", round(percentVar[1] * 100), "% variance")) + 
      ylab(paste0("PC2: ", round(percentVar[2] * 100), "% variance")) +
      scale_colour_manual(values = mycols) + 
      scale_x_continuous(limits = xlim) + 
      scale_y_continuous(limits = ylim) + 
      facet_grid(.~Cohort) +
      guides(size = FALSE, alpha = FALSE) + theme_bw(base_size = 12)

  show(p)
  #grid.arrange( grobs = gl, ncol = 2)
  print(summary(lm(PC2 ~ Cohort, data = pd)))
  print(var.test(PC2 ~ Cohort, data = pd))
  
  pd
}

```

## PCA of Combined Discovery and Replicatiion Cohorts 

Do PCA on top 5000 genes by expression (using both cohorts). Also show stats for lm (t-test) and var.test (F-test of variance) on PC2

The lower plots are from the same PCA analysis, filtered to display data points from only one cohort

```{r pca1}
pca1 <- pca_fn(voomed_data$E)

```


## Apply removeBatchEffect 

limma::removeBatchEffect fits a linear model with design matrix ~Cohort to the expression matrix E , and returns E - beta %*% design_matrix. It brings the two cohorts as much into register as possible, so that there is no longer any significant difference in the mean of PC2 between Discovery and Replication cohorts, and also much reduces the apparent difference in variance between them. 

This strongly suggests that, in performing Differential Expression analysis on the combined D + R cohorts, a main effect for the Cohort should be included. 


```{r pca2}
voomed_data2 <- removeBatchEffect(voomed_data$E, batch = keep_samp$Cohort)
pca2 <- pca_fn(voomed_data2)


```

